Block bootstrap consistency\\under weak assumptions

### 2015-12-31 Paper.tex
[#3] Resolve variance issue raised by referee
This commit removes corollary 1 and tries to resolve some issues
with the covariance process that one of the referees found confusing.

Theorem~\ref{main-bootstrap-clt} can be further extended to give an
\fclt\ using arguments from \citet{JoD:00b}. We show in
Theorem~\ref{main-bootstrap-invariance-theorem} that the partial sum
of the bootstrapped process obeys an \fclt\ and can be used to derive
critical values for other test statistics under the same assumptions
as Theorem~\ref{main-bootstrap-clt}.

\begin{thm}\label{main-bootstrap-invariance-theorem}
  Suppose that the conditions of Theorem~\ref{main-bootstrap-clt}
  hold, let $W$ be standard Brownian Motion, and define
  \begin{equation}
    \label{eq:6}
    W_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \sigma_{n}^{*}
  \end{equation}
  and
  \begin{equation}
    \label{eq:7}
    \hat{W}_n^{*}(\gamma) = \tfrac{1}{\sqrt{n}}
    \sum_{t=1}^{\lfloor \gamma n \rfloor} (X^*_{nt} - \mu_n^{*}) / \hat\sigma_n^{*}.
  \end{equation}
  Then $\pr^{*}[\dist(W_n^{*}, W) > \delta] \to 0$ i.p.\ and
  $\pr^{*}[\dist(\hat{W}_n^{*}, W) > \delta] \to 0$ i.p.\ for any
  positive $\delta$ and distance $\dist$ that metricizes weak
  convergence.
\end{thm}

Obviously, Theorem~\ref{main-bootstrap-invariance-theorem}, has
natural corollaries that allow it to be used to approximate the
behavior of partial sums of the original series.  But applying this
result requires stronger assumptions than
Theorem~\ref{main-bootstrap-clt}. The bootstrapped processes are
(essentially) homoskedastic, so they are unable to match patterns of
heteroskedasticity in the original series and we can not estimate the
covariance process of the original partial sum in the bootstrap. This
necessitates a separate consistent estimator of the covariance process, which
we did not need for the sample mean.  Other methods, such as the Local
Block Bootstrap \citep{PaP:02,DPP:03}, may be able to capture this
additional heterogeneity with the bootstrap alone, but we do not
pursue that possibility further.

\begin{cor}
  Suppose the conditions of Theorem~\ref{main-bootstrap-clt} hold and
  that $\sup_{t=1,\dots,n}|\mu_{nt} - \bar \mu_n| = o(1/\sqrt{n})$. Define
  \begin{equation}
    \label{eq:8}
    \sigma_n^2(\gamma) = \tfrac{1}{n}
    \sum_{s,t=1}^{\lfloor \gamma n \rfloor} \cov(X_{ns}, X_{nt}),
  \end{equation}
  let $\hat\sigma_n^2(\cdot)$ be an estimator of $\sigma_n^2(\cdot)$
  such that $\sup_{\gamma} |\sigma_n^2(\gamma) /
  \hat{\sigma}_n^2(\gamma) - 1| \to^p 0$, and let
  \begin{equation}
    \label{eq:9}
    \hat{W}_n(\gamma) = \tfrac{1}{\sqrt{n}} \sum_{t=1}^{\lfloor \gamma
      n \rfloor} (X_{nt} - \bar\mu_n) / \hat{\sigma}_{n}(\gamma).
  \end{equation}
  Then $\pr^{*}[\dist(W_n^{*}, \hat{W}_n) > \delta] \to^p 0$ and
  $\pr^{*}[\dist(\hat{W}_n^{*}, \hat{W}_n) > \delta] \to^p 0$ for any
  positive $\delta$ and any distance $\dist$ that metricizes weak
  convergence.
\end{cor}

@@

\subsection*{Proof of Theorem~\ref{main-bootstrap-invariance-theorem}}
\newcommand{\gn}{{\lfloor \gamma n \rfloor}}
\newcommand{\gJ}{{\lfloor \gamma J_n \rfloor}}
\newcommand{\lastK}{{K_{n, \gJ}}}
Define $W_n^{\prime*}$ as
\begin{equation*}
  W^{\prime*}_n(\gamma)
  = \tfrac{1}{\sqrt{n}} \sum_{t=1}^\gn (X_{nt}^* - \bar X_{n})/\sigma
  = \sum_{j=1}^\gJ \Zb /\sigma + Z_n^*(\lastK, \gn - \lastK) / \sigma.
\end{equation*}
Lemma~\ref{bootstrap-variance-convergence-lemma} ensures that
$\sigma_n^{*2}$ and $\hat\sigma_n^{*2}$ both converge in probability to
$\sigma^2$, so it suffices to prove that $\pr^*[\dist(W_n^{\prime*}, W) >
\delta)] \to^p 0$; we can also assume without loss of generality that
$\sigma^2 = 1$. Moreover,
Lemma~\ref{equivalent-convergence-from-lie-lemma} ensures that it
suffices to prove unconditional convergence, so we will establish
$\pr[\dist(W_n^{\prime *}, W) > \delta] \to 0$.

Theorem~\ref{main-bootstrap-clt} implies that, for any fixed $\gamma$,
$W_n^{\prime *}(\gamma)$ is asymptotically normal and, as a
partial sum of an \mds, $W_n^{\prime *}$ has asymptotically independent
increments. To see this, choose $\gamma, \gamma' \in [0,1]$ and
$\delta, \delta' > 0$ so that $\delta + \gamma \leq \gamma'$. Since
the increments of $W_n^{\prime *}$ are uncorrelated conditional on
$X_{1n},\dots,X_{nn}$, and $\Ms$, we have
\begin{equation*}
    \E\big[ (W_n^{\prime *}(\delta' + \gamma') - W_n^{\prime *}(\gamma'))
  (W_n^{\prime *}(\delta + \gamma) - W_n^{\prime *}(\gamma)) \big] = 0
\end{equation*}
for large enough $n$ if $\gamma' > \gamma + \delta$. If $\gamma' =
\gamma + \delta$ then
\begin{multline*}
  \E\big[(W_n^{\prime *}(\delta' + \gamma') - W_n^{\prime *}(\gamma'))
  (W_n^{\prime *}(\delta + \gamma) - W_n^{\prime *}(\gamma)) \big] = \\
  \E \Em^*\big\{- Z_n^*(\lfloor \gamma' n \rfloor, K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)
  Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))\big\}.
\end{multline*}
But this second quantity can be bounded:
\begin{align*}
  \E \Em^*\big\{-& Z_n^*(\lfloor \gamma' n \rfloor,
  K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)
  Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))\big\} \\
  &\leq \big\{\E \Em\big[Z_n^*(\lfloor \gamma' n \rfloor, K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)\big]^2 \\
  & \quad \times \big\{\E \Em Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))^2\big\}^{1/2} \\
  &\leq C \E M_{n, \lfloor (\gamma + \delta) J_n \rfloor} / n
\end{align*}
for some constant $C$ by
Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma}. This term
converges to zero by Lemma~\ref{block-length-convergence-lemma}.

Consequently, as in \cite{JoD:00b}, the result follows from stochastic
equicontinuity \citep[Theorems 15.4 and 15.5]{Bil:68} namely that
\begin{equation*}
  \lim_{\delta \to 0} \limsup_{n \to \infty} \pr[\sup_{\gamma \in [0,1]}
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
    | W_n^{\prime *}(\gamma) - W_n^{\prime *}(\gamma') | > \epsilon] = 0
\end{equation*}
for any positive $\epsilon$.  Fix $\delta> 0$ such that $D = 2/\delta$
is a positive integer and let $\gamma_d = d/D$ for $d =
0,\dots,D$. Then, mimicking the argument in \cite{JoD:00b},
\begin{align*}
  \pr[\sup_{\gamma \in [0,1]} &
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
  | W_n^{\prime *}(\gamma) - W_n^{\prime *}(\gamma') | > \epsilon] \\
  &\leq \pr[ \sup_{d = 1,\dots,D} \sup_{\gamma \in [0,\delta]}
  | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) | > \epsilon/2] \\
  &\leq  (4/\epsilon^2) \sum_{d=1}^D \E[\sup_{\gamma \in [0,\delta]}
  | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2 \\
  &\qquad\qquad\qquad \times \ind\{\sup_{\gamma \in [0,\delta]}
  | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2 > \epsilon^2/4\}] \\
  &\leq  (4/\epsilon^2) \max_{d=1,\dots,D} \E[\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2 \\
  &\qquad\qquad\qquad \times \ind\{\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2 > \epsilon^2/4\delta\}].
\end{align*}
Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma} implies that
\begin{equation*}
  \sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2
\end{equation*}
is uniformly integrable, so there exists a finite and positive
function $B$ such that $B(x) \to 0$ as $x \to \infty$ and
\begin{multline*}
  \E[\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2
  \ind\{\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{\prime *}(\gamma + \gamma_d) - W_n^{\prime *}(\gamma_d) |^2 > x\}] \\
  \leq B(x)
\end{multline*}
for all $d$ and $\delta$ and all large enough $n$.

As a result,
\begin{align*}
  \lim_{\delta \to 0} \limsup_{n\to\infty} \pr[\sup_{\gamma \in [0,1]}
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
  | W_n^{\prime *}(\gamma) - W_n^{\prime *}(\gamma') | > \epsilon] & \leq
  \lim_{\delta \to 0} (4/\epsilon^2)
  B(\epsilon^2/4\delta) \\
  &= 0,
\end{align*}
completing the proof.\qed

>>>>

Theorem~\ref{main-bootstrap-clt} can also be extended to give an \fclt\
using arguments from \citet{JoD:00b}. We show in
Theorem~\ref{main-bootstrap-invariance-theorem} that the partial sum
of the bootstrapped process obeys an \fclt\ and can be used to derive
critical values for other test statistics under the same assumptions
as Theorem~\ref{main-bootstrap-clt}. For this result, define the
following partial sums,
\begin{align*}
W_n(\gamma) &=
\tfrac{1}{\sqrt{n}} \sum_{t=1}^{\lfloor \gamma n \rfloor}
(X_{nt} - \bar\mu_n) &&\text{and}&
W_n^{*}(\gamma) &= \tfrac{1}{\sqrt{n}}
\sum_{t=1}^{\lfloor \gamma n \rfloor}
(X^*_{nt} - \mu_n^{*}).
\end{align*}
Also let $W$ denote standard Brownian Motion and $\sigma W$ denote
Brownian Motion scaled by the constant $\sigma$.

\begin{thm}\label{main-bootstrap-invariance-theorem}
Suppose that the conditions of Theorem~\ref{main-bootstrap-clt} hold
and let $\dist$ be any distance function that metricizes weak
convergence. Then
\begin{equation}\label{eq:7}
\pr^{*}[\dist(W_n^{*}, \sigma W) > \delta] \to^p 0
\end{equation}
for any positive  $\delta$.
If, in addition, $\sup_{t=1,\dots,n}|\mu_{nt} -
\bar \mu_n| = o(1/\sqrt{n})$ and
\begin{equation}\label{eq:4}
n^{-1} \sum_{s,t=1}^{\lfloor \gamma n \rfloor} \cov(X_{ns}, X_{nt}) \to \sigma^2 \gamma
\end{equation}
for all $\gamma \in [0,1]$, then
\begin{equation}\label{eq:8}
\pr[\dist(W_n, \sigma W) > \delta] \to 0
\end{equation}
for any positive $\delta$.
\end{thm}
If both~\eqref{eq:7} and~\eqref{eq:8} hold then the bootstrap can be
used to approximate the distribution of partial sums.
Note that Theorem~\ref{main-bootstrap-invariance-theorem} imposes
stronger assumptions for the original process
than for the bootstrapped process. Without~\eqref{eq:4}, the
partial sum of the original observations converges to a transformed
Brownian Motion with a different covariance process. The bootstrapped
partial sum, on the other hand, always converges to standard (but
potentially scaled) Brownian Motion because the resampling strategies
ensure that the bootstrapped process is globally covariance stationary.

If the original process does not satisfy~\eqref{eq:4}, it would be
necessary to normalize $W_n$ with a uniformly
consistent estimator of the true covariance process of the series to
make use of these results. This would be the case if the variance
permanently changes partway through the series, for example.  Other
methods, such as the Local Block Bootstrap \citep{PaP:02,DPP:03}, may
be able to capture this additional heterogeneity with the bootstrap
alone, but we do not pursue that possibility further.

@@

\subsection*{Proof of Theorem~\ref{main-bootstrap-invariance-theorem}}
\newcommand{\gn}{{\lfloor \gamma n \rfloor}}
\newcommand{\gJ}{{\lfloor \gamma J_n \rfloor}}
\newcommand{\lastK}{{K_{n, \gJ}}}

We will only present proofs for the bootstrap results, since
Theorem~3.1 of \cite{JoD:00b} establishes the result for the partial
sums of the original process.  Theorem~\ref{main-bootstrap-clt}
implies that, for any fixed $\gamma$, $W_n^{*}(\gamma)$ is
asymptotically normal with limiting variance $\gamma \sigma^2$, so we
can assume that $\sigma^2 = 1$ without loss of generality. Moreover,
Lemma~\ref{equivalent-convergence-from-lie-lemma} shows that it is
sufficient to prove unconditional convergence, so we will establish
$\pr^*[\dist(W_n^{*}, W) > \delta)] \to^p 0$. As in \cite{JoD:00b},
this will hold if we show that $W_n^{*}$ has asymptotically
independent increments and stochastic equicontinuity, namely
\begin{equation}\label{eq:9}
  \lim_{\delta \to 0} \limsup_{n \to \infty} \pr[\sup_{\gamma \in [0,1]}
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
    | W_n^{*}(\gamma) - W_n^{*}(\gamma') | > \epsilon] = 0
\end{equation}
for any positive $\epsilon$.

First observe that we can write $W_n^{*}$ as
\begin{equation*}
  W^{*}_n(\gamma)
  = \tfrac{1}{\sqrt{n}} \sum_{t=1}^\gn (X_{nt}^* - \bar X_{n})
  = \sum_{j=1}^\gJ \Zb + Z_n^*(\lastK, \gn - \lastK).
\end{equation*}
To show that the increments of this process are asymptotically
independent, choose $\gamma, \gamma' \in [0,1]$ and
$\delta, \delta' > 0$ so that $\delta + \gamma \leq \gamma'$. Since
the blocks of $W_n^{*}$ are conditionally uncorrelated given
$X_{1n},\dots,X_{nn}$, and $\Ms$, we have
\begin{equation*}
  \E\big[ (W_n^{*}(\delta' + \gamma') - W_n^{*}(\gamma'))
  (W_n^{*}(\delta + \gamma) - W_n^{*}(\gamma)) \big] = 0
\end{equation*}
for large enough $n$ if $\gamma' > \gamma + \delta$. If $\gamma' =
\gamma + \delta$ then
\begin{multline*}
  \E\big[(W_n^{*}(\delta' + \gamma') - W_n^{*}(\gamma'))
  (W_n^{*}(\delta + \gamma) - W_n^{*}(\gamma)) \big] = \\
  \E \Em^*\big\{- Z_n^*(\lfloor \gamma' n \rfloor, K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)
  Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))\big\}.
\end{multline*}
But this second quantity can be bounded:
\begin{align*}
  \E \Em^*\big\{-& Z_n^*(\lfloor \gamma' n \rfloor,
  K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)
  Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))\big\} \\
  &\leq \big\{\E \Em\big[Z_n^*(\lfloor \gamma' n \rfloor, K_{n, \lceil \gamma' J_n \rceil }- \lfloor \gamma' n \rfloor)\big]^2 \\
  & \quad \times \big\{\E \Em Z_n^*(K_{n,\lfloor (\gamma + \delta) J_n \rfloor},
  \lfloor (\gamma + \delta) n \rfloor - K_{n,\lfloor (\gamma + \delta) J_n \rfloor}))^2\big\}^{1/2} \\
  &\leq C \E M_{n, \lfloor (\gamma + \delta) J_n \rfloor} / n
\end{align*}
for some constant $C$ by
Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma}. This term
converges to zero by Lemma~\ref{block-length-convergence-lemma}.

For~\eqref{eq:9}, fix $\delta> 0$ such that $D = 2/\delta$ is a
positive integer and let $\gamma_d = d/D$ for $d =
0,\dots,D$. Mimicking the argument in \cite{JoD:00b} gives the bounds
\begin{align*}
  \pr[\sup_{\gamma \in [0,1]} &
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
  | W_n^{*}(\gamma) - W_n^{*}(\gamma') | > \epsilon] \\
  &\leq \pr[ \sup_{d = 1,\dots,D} \sup_{\gamma \in [0,\delta]}
  | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) | > \epsilon/2] \\
  &\leq  (4/\epsilon^2) \sum_{d=1}^D \E[\sup_{\gamma \in [0,\delta]}
  | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 \\
  &\qquad\qquad\qquad \times \ind\{\sup_{\gamma \in [0,\delta]}
  | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 > \epsilon^2/4\}] \\
  &\leq  (4/\epsilon^2) \max_{d=1,\dots,D} \E[\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 \\
  &\qquad\qquad\qquad \times \ind\{\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 > \epsilon^2/4\delta\}].
\end{align*}
Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma} implies that
\begin{equation*}
  \sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2
\end{equation*}
is uniformly integrable, so an argument similar to that used in the
proof of Theorem~\ref{main-bootstrap-clt} shows that there exists a
finite and monotone function $B$ such that $B(x) \to 0$ as
$x \to \infty$ and
\begin{multline*}
  \E[\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 \times \\
  \ind\{\sup_{\gamma \in [0,\delta]}
  (1/\delta) | W_n^{*}(\gamma + \gamma_d) - W_n^{*}(\gamma_d) |^2 > x\}]
  \leq B(x)
\end{multline*}
for all $d$ and $\delta$ and all large enough $n$.

As a result,
\begin{align*}
  \lim_{\delta \to 0} \limsup_{n\to\infty} \pr[\sup_{\gamma \in [0,1]}
  \sup_{\gamma' \in [\gamma - \delta, \gamma + \delta]}
  | W_n^{*}(\gamma) - W_n^{*}(\gamma') | > \epsilon] & \leq
  \lim_{\delta \to 0} (4/\epsilon^2)
  B(\epsilon^2/4\delta) \\
  &= 0,
\end{align*}
completing the proof.\qed
